---
title: "Introduction to R, part 4d"
author: "Steve Simon"
date: "Sunday, August 14, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to convert a longitudinal data set from the tall and thin format to a short and fat format.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

# Part 4d. another reshape2 example
 
Erase any stray data.

```{r clean-up}
rm(list=ls())
load("part4.RData")
```

## Converting from tall and thin to short and fat

Now let's read in a tall and thin format and convert it to a
short and wide format.

This data set is described at 

--> http://www.statsci.org/data/general/cholest.html

```{r chol-read}
fi <- "http://www.statsci.org/data/general/cholestg.txt"
ch <- read.table(fi,header=TRUE)
head(ch)
tail(ch)
summary(ch)
```

The summary is not too helpful for the first three variables, so 
try the table command instead.

```{r chol-table}
table(ch$patient,useNA="always")
table(ch$group,useNA="always")
table(ch$day,useNA="always")
```

It's still a bit confusing. Why do the first 28 subjects have three
measurements and the last 30 have only one measurement.

```{r chol-crosstabs}
table(ch$patient,ch$day,useNA="always")
table(ch$group,ch$day,useNA="always")
```

Okay, now it's a bit more clear. Each subject has measurements on
three days (2,4,14) for group one, and measurements on an unspecified
day (NA) for group 2.

Let's look at the tedious and the faster way to convert from the tall
and thin to the short and fat format.

The missing value for day is going to cause a lot of grief, and it is
simpler here to change it to a special code, 99.

```{r chol-reshape-1}
ch$day[is.na(ch$day)] <- 99

sb02 <- ch[ch$day==2,c(1,4)]
names(sb02)[2] <- "cholest02"
head(sb02)

sb04 <- ch[ch$day==4,c(1,4)]
names(sb04)[2] <- "cholest04"
head(sb04)

sb14 <- ch[ch$day==14,c(1,4)]
names(sb14)[2] <- "cholest14"
head(sb14)

sb99 <- ch[ch$day==99,c(1,4)]
names(sb99)[2] <- "cholest99"
head(sb99)

me1 <- merge(sb02,sb04,by="patient")
head(me1)
tail(me1)

me2 <- merge(me1,sb14,by="patient")
head(me2)
tail(me2)

me3 <- merge(me2,sb99,by="patient")
head(me3)
tail(me3)
```

Oops! What happened here is that no matches were found and the default
in the merge function is to only include in the merge those rows that
are found in both data sets. This is known as an inner join in database
parlance. You can use the all.x and all.y arguments to use a left join,
right join, or outer join instead.

```{r chol-reshape2}
me3 <- merge(me2,sb99,by="patient",all.x=TRUE,all.y=TRUE)
head(me3)
tail(me3)
```

```{r chol-reshape-3}
library("tidyr")
ch1 <- spread(ch, day, cholest)
head(ch1)
tail(ch1)
cor(ch1[,c("2", "4", "14")])
```

Oops! The missing values hurt us here. Review the help file for cor.

```{r cholest-cor}
cor(ch1[,c("2", "4", "14")],use="complete")
```

## A data set on back pain with two matched control groups

This data set is described in
--> http://www.statsci.org/data/oz/backpain.html

Subjects were matched by age, height, and weight. How well were they matched?

```{r read-match}
fn <- "http://www.statsci.org/data/oz/backpain.txt"
ma <- read.table(fn,header=TRUE)
ma <- ma[,1:6]
head(ma)
tail(ma)
```

## On your own

Convert this to a short and fat format. Let age1 represent the age of
the subject in the pain group, age2 represent the matched subject
in the nopain group, and age3 represent the matched subject in the
sedentary group. Redefine height and weight similarly.

Plot age1 versus age2 and age1 versus age3 to evaluate how closely
the subjects are matched on age. Repeat for height and weight.

## If there is extra time

Here is a very complex data set on stroke.
--> http://www.statsci.org/data/oz/stroke.html

```{read-stroke}
fi <- "http://www.statsci.org/data/oz/stroke.txt"
st <- read.table(fi,header=TRUE)
head(st)
tail(st)
```

Save everything

```{r save-everything}
save.image("part4.RData")
```

## There is no special assignment to be handed in today.

If you have extra time, work on your own data set.
