---
title: "Introduction to R, part 4c"
author: "Steve Simon"
date: "Sunday, August 14, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to use the reshape2 package to convert a longitudinal data set from the short and fat format to the tall and thin format.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

Erase any stray data and load the data from part 4b.

```{r read-part-4b-data}
rm(list=ls())
load(file="part4.RData")
```

The reshape2 library is very complex, but for what you need to do here today, you only need one function: melt.

```{r faster-1}
library("reshape2")
teb <- melt(te,id.vars=c("dish","dose"))
head(teb)
tail(teb)
```

It takes a bit of investigation, but the warning messages refer to
the day3 and day9 values which are character strings. You can fix this by
droping those columns.

```{r faster-2}
tec <- melt(te[,c(-5,-11)],id.vars=c("dish","dose"))
head(tec)
tail(tec)
```

We're not done quite yet. The day of the week is hidden in the variable
called "variable" and we need to keep only the fourth (and maybe fifth)
characters.

```{r substr-1}
substr(tec$variable,4,5)
```

Almost. We need to convert from character to numeric.

```{r substr-2}
tec$day <- as.numeric(substr(tec$variable,4,5))
head(tec)
tail(tec)
```

I'm not too thrilled with the non-descriptive name, value, but
I'm also too lazy here to change it or to drop the now unneeded
column, variable.

Now to plot the data.

```{r plot-termites}
plot(tec$day,tec$value)
```

Let's connect the data for individual subjects.

```{r line-termites}
plot(tec$day,tec$value)
for (i in 1:8) {
  sb <- tec$dish==i&tec$dose==5
  lines(tec$day[sb],tec$value[sb],col="darkred")
  sb <- tec$dish==i&tec$dose==10
  lines(tec$day[sb],tec$value[sb],col="darkgreen")
}
```

If you look at the graph, there are a couple of green lines, which
represent dose=10, but which have many more alive at day 15 than even
the dose=5 values. If you look at the original data set, these are
rows 9 and 10, corresponding to the first two dishes in the dose=10
group. This may be okay, but you should check with the people who 
collected the data, if you can.

## A data set on depression after an earthquake (short and fat format)

This data set is described in 
--> http://www.statsci.org/data/general/lomaprie.html


```{r read-earthquake}
fn <- "http://www.statsci.org/data/general/lomaprie.txt"
qu <- read.table(fn,header=TRUE)
qu
```

## On your own

Convert this to a tall and thin format. Draw boxplots.

## A data set on energy consumption (short and fat format)

This data set is described at

--> http://www.statsci.org/data/general/energy.html

```{r read-energy}
fi <- "http://www.statsci.org/data/general/energy.txt"
en <- read.table(fi,header=TRUE)
head(en)
tail(en)
```

## Save all my work

```{r save-image}
save.image("part4.RData")
```
